---
output: github_document
---

# ancemb

<!-- badges: start -->
[![R-CMD-check](https://github.com/alrobles/ancemb/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/alrobles/ancemb/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/alrobles/ancemb/actions/workflows/pkgdown.yaml/badge.svg)](https://alrobles.github.io/ancemb/)
[![License: GPL (>= 3)](https://img.shields.io/badge/license-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

**Bayesian ancestral reconstruction of protein language-model embeddings.**

`ancemb` treats the embedding produced by a protein language model (pLM)
as a set of continuous traits evolving along a phylogeny. It reconstructs
ancestral embeddings at internal nodes under Brownian Motion (BM) and
Ornstein–Uhlenbeck (OU) processes, fitted with pre-compiled
[Stan](https://mc-stan.org) models through the
[`instantiate`](https://wlandau.github.io/instantiate/) package.

## What we are doing

Protein language models compress each sequence into a high-dimensional
vector that encodes structure and function. A natural question follows:
do those coordinates behave like phylogenetic traits — and if so, where
does ancestral information stop being recoverable from embedding
geometry?

This package is the analysis interface of a research program that
decomposes that question into three testable hypotheses:

- **T1 — trait evolution.** Embedding dimensions carry components that
  evolve like continuous traits on the tree (measurable as Blomberg's *K*,
  Pagel's *λ*, predictive density on held-out clades).
- **T2 — trait reconstruction.** Phylogenetic models (BM/OU) reconstruct
  ancestral embeddings better than phylogeny-free baselines.
- **T3 — invertibility.** Closeness in embedding space implies recovery of
  the ancestral *sequence*.

Keeping these separate is deliberate: a failure of T3 says nothing about
T2 unless the same nodes, pools, and metrics are held fixed — which is
what this package enforces.

## What we intend

`ancemb` aims to be the single, tested entry point for ancestral embedding
inference: data preparation, Stan posterior fits, pure-R reference
implementations, an optional Python (reticulate) backend for embedding
arrays, and manifest-driven metrics that make results comparable across
families, representations, and models. The long-term goal is a calibrated
predictive account of embedding evolution — including an honest map of
where it fails.

## What has been achieved so far

- **Validated estimator.** On Gaussian-trait simulations where the truth
  is known by construction, the BM/OU fits recover the simulated
  ancestors. On real protein families, Bayesian posterior means agree
  with `phytools::fastAnc` ML reconstructions to cosine ≈ 0.99.
- **Partial phylogenetic signal.** Across a curated multi-family database
  (trees + alignments + deposited and experimentally resurrected
  ancestors), ~10.7% of embedding dimensions show significant Blomberg's
  *K* after multiple-testing correction; ancestral reconstruction beats
  phylogeny-free baselines in shallow families and degrades with depth.
- **An honest limit.** Sequence-level recovery by ranking candidates in
  embedding space does *not* beat classical MAP reconstruction — in a
  340-node benchmark, MAP wins as soon as real headroom exists
  (P(d > 0) ≈ 0.19). A linear probe recovers ancestral sequence
  information above all phylogeny-free baselines, so the information is
  present in the embeddings; the failing component is the
  cosine-ranking/decoder step. Measuring that boundary — not claiming
  superiority — is the contribution.

A manuscript describing the full benchmark is in preparation.

## Installation

`ancemb` requires a working
[CmdStan](https://mc-stan.org/users/interfaces/cmdstan) installation.
Install CmdStan from R with:

```r
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()
```

Then install `ancemb` from GitHub:

```r
# install.packages("remotes")
remotes::install_github("alrobles/ancemb")
```

## Quick start

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r setup}
library(ancemb)
set.seed(42)
```

Prepare a small tree and some observed tip embeddings:

```{r data}
tree <- ape::rtree(8)
tree$edge.length <- tree$edge.length / max(ape::node.depth.edgelength(tree))

z_obs <- matrix(stats::rnorm(8 * 4, sd = 0.5), nrow = 8)
rownames(z_obs) <- tree$tip.label
```

Build the Stan data list and fit the Brownian Motion model:

```{r fit, eval = instantiate::stan_cmdstan_exists()}
stan_data <- prepare_embedding_data(tree, z_obs, obs_sigma = 0.1)
fit <- fit_embedding_bm(stan_data, chains = 2, iter_warmup = 300,
                        iter_sampling = 300, refresh = 0)
```

Extract ancestral states and compare the inferred root with `phytools::fastAnc`:

```{r extract, eval = instantiate::stan_cmdstan_exists()}
anc <- extract_ancestral(fit, data = stan_data, tree = tree)
root_node <- which(anc$node == stan_data$S + 1)

ml_root <- phytools::fastAnc(tree, z_obs[, 1])[1]
cat("ML root (dim 1):", ml_root, "\n")
cat("Bayes root mean (dim 1):", anc$mean[root_node, "dim1"], "\n")
```

## Reference implementations

Every exported Stan wrapper is paired with a pure-R reference function
(`*_r`) that reproduces the same mathematics line by line:
`simulate_bm_embedding_r()`, `simulate_ou_embedding_r()`,
`reconstruct_ancestral_bm_r()`, `reconstruct_ancestral_ou_r()` and
`reconstruct_ancestral_ml_r()`. The test suite cross-checks Stan
posterior means against both these references and `phytools::fastAnc()`.

## Python backend

An optional reticulate adapter (`configure_python_backend()`,
`python_backend_status()`, `load_embedding_array()`) reads NumPy `NPZ`
embedding archives directly, so the R pipeline can consume pLM outputs
without intermediate conversions.

## Roadmap

- Representation geometry: whitening, PCA, probabilistic PCA and
  invertible autoencoders as pre-defined analysis arms.
- Held-out-clade (LOCO) predictive benchmark against sequence-only
  controls, with pre-registered admission gates.
- Decoders beyond cosine ranking over fixed candidate pools.

## Citation

`ancemb` is research software under active development. If you use it,
please cite the GitHub repository; a manuscript citation will be added
when available.

## Code of Conduct

Please note that this project is released with a [Contributor Code of
Conduct](https://contributor-covenant.org/version/2/1/code_of_conduct.html).
By participating in this project you agree to abide by its terms.
