---
title: "ancemb: Bayesian Ancestral Protein Embeddings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ancemb: Bayesian Ancestral Protein Embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`ancemb` provides a small set of functions to reconstruct ancestral protein
embeddings from extant (tip) embeddings and a phylogenetic tree. The package
uses pre-compiled Stan models for Brownian Motion (BM) and
Ornstein-Uhlenbeck (OU) evolution of continuous embedding dimensions.

## Setup

```{r setup}
library(ancemb)
set.seed(42)
```

## Simulated data

We create a small tree and a synthetic set of tip embeddings:

```{r data, eval = requireNamespace("instantiate", quietly = TRUE) && instantiate::stan_cmdstan_exists()}
tree <- ape::rtree(8)
tree$edge.length <- tree$edge.length / max(ape::node.depth.edgelength(tree))

z_obs <- matrix(stats::rnorm(8 * 4, sd = 0.5), nrow = 8)
rownames(z_obs) <- tree$tip.label
```

## Fitting the BM model

```{r fit, eval = requireNamespace("instantiate", quietly = TRUE) && instantiate::stan_cmdstan_exists()}
stan_data <- prepare_embedding_data(tree, z_obs, obs_sigma = 0.1)
fit <- fit_embedding_bm(stan_data, chains = 1, iter_warmup = 100,
                        iter_sampling = 100, refresh = 0)
```

## Extracting ancestral states

`extract_ancestral()` computes posterior summaries for all internal nodes:

```{r extract, eval = requireNamespace("instantiate", quietly = TRUE) && instantiate::stan_cmdstan_exists()}
anc <- extract_ancestral(fit, data = stan_data, tree = tree)
str(anc)
```

## Comparing model families

```{r compare, eval = requireNamespace("instantiate", quietly = TRUE) && instantiate::stan_cmdstan_exists()}
cmp <- run_bayesian_comparison(stan_data, tree, chains = 1, iter_warmup = 100,
                               iter_sampling = 100, refresh = 0)
cmp$root_comparison
```

## Pure-R reference

Every exported function is paired with a pure-R reference implementation
(`*_r`) that follows the math line-by-line. For example,
`reconstruct_ancestral_bm_r()` matches `phytools::fastAnc()` for noise-free
BM data, which is used in the package test suite as the cross-language QA
benchmark.
